---
title: "ISIÔÇôT2D Genetic Association Browser"
output:
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    theme:
      version: 4
      bootswatch: flatly
    css: www/custom.css
    favicon: ""
runtime: shiny
---

```{r global, include=FALSE}
library(shiny)
library(flexdashboard)
library(plotly)
library(DT)
library(dplyr)
library(tidyr)
library(readr)
library(purrr)
library(here)
library(scales)

`%||%` <- function(a, b) if (!is.null(a) && length(a) > 0) a else b

source(here::here("R/load_data.R"))
source(here::here("R/plots.R"))

# ÔöÇÔöÇ Load data once at startup ÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇ
data_list  <- load_combined_data(here::here("data"))
dat_bmi    <- data_list$bmi
dat_no_bmi <- data_list$no_bmi

# Pre-join for the comparison views
dat_joined <- dplyr::inner_join(
  dat_bmi    %>% dplyr::rename(beta_bmi = beta, pval_bmi = pvalue),
  dat_no_bmi %>% dplyr::rename(beta_no_bmi = beta, pval_no_bmi = pvalue),
  by = c("nearest_gene","index_variant","t_2_d_risk_allele",
         "gene_snp_ra","is_index","index_group")
)

all_variants <- sort(unique(c(dat_bmi$gene_snp_ra, dat_no_bmi$gene_snp_ra)))
```


Sidebar {.sidebar data-width=265}
=====================================

**Analysis & Filters**

```{r}
radioButtons("analysis", "Analysis",
  choices  = c("BMI-Adjusted" = "bmi", "BMI-Unadjusted" = "no_bmi"),
  selected = "bmi",
  inline   = FALSE
)

hr()

textInput("search", label = "­ƒöì Gene name / rsID",
          placeholder = "e.g.  KCNJ11  or  rs5219")

checkboxGroupInput("groups", "Index Groups",
  choices  = c("Fasting", "OGTT,0-120", "OGTT,0-30-120"),
  selected = c("Fasting", "OGTT,0-120", "OGTT,0-30-120")
)

hr()

sliderInput("pval",     "Max p-value",   min = 0, max = 1,    value = 1,    step = 0.05)
sliderInput("beta_abs", "Min |╬▓|",       min = 0, max = 0.03, value = 0,    step = 0.001)

hr()

actionButton("reset_btn", "Ôƒ│  Reset filters",
             class = "btn btn-outline-secondary btn-sm w-100")
```

---

```{r}
renderUI({
  d <- fd()
  tagList(
    tags$div(style = "font-size:0.8rem; color:#555; line-height:1.9",
      tags$b("Active dataset"), tags$br(),
      tags$span(dplyr::n_distinct(d$gene_snp_ra),  " variants"),    tags$br(),
      tags$span(dplyr::n_distinct(d$nearest_gene), " unique genes"), tags$br(),
      tags$span(sum(d$pvalue < 0.05),  " sig. assoc. (p<0.05)"),    tags$br(),
      tags$span(sum(d$pvalue < 0.001), " sig. assoc. (p<0.001)")
    )
  )
})
```


```{r}
# ÔöÇÔöÇÔöÇ Shared reactive: filtered data for the active analysis ÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇÔöÇ
fd <- reactive({
  df <- if (input$analysis == "bmi") dat_bmi else dat_no_bmi
  q  <- tolower(trimws(input$search %||% ""))
  grps <- input$groups %||% c("Fasting", "OGTT,0-120", "OGTT,0-30-120")

  df %>%
    dplyr::filter(
      index_group %in% grps,
      pvalue      <= (input$pval     %||% 1),
      abs(beta)   >= (input$beta_abs %||% 0)
    ) %>%
    dplyr::filter(
      if (nchar(q) > 0)
        grepl(q, tolower(nearest_gene),  fixed = TRUE) |
        grepl(q, tolower(index_variant), fixed = TRUE) |
        grepl(q, tolower(gene_snp_ra),   fixed = TRUE)
      else
        rep(TRUE, dplyr::n())
    )
})

# Filtered joined dataset (both analyses, same gene/variant filter)
fj <- reactive({
  q    <- tolower(trimws(input$search %||% ""))
  grps <- input$groups %||% c("Fasting", "OGTT,0-120", "OGTT,0-30-120")

  dat_joined %>%
    dplyr::filter(index_group %in% grps) %>%
    dplyr::filter(
      if (nchar(q) > 0)
        grepl(q, tolower(nearest_gene),  fixed = TRUE) |
        grepl(q, tolower(index_variant), fixed = TRUE) |
        grepl(q, tolower(gene_snp_ra),   fixed = TRUE)
      else
        rep(TRUE, dplyr::n())
    )
})

# Reset all controls to defaults
observeEvent(input$reset_btn, {
  updateRadioButtons(    session, "analysis", selected = "bmi")
  updateTextInput(       session, "search",   value    = "")
  updateCheckboxGroupInput(session, "groups", selected = c("Fasting","OGTT,0-120","OGTT,0-30-120"))
  updateSliderInput(     session, "pval",     value    = 1)
  updateSliderInput(     session, "beta_abs", value    = 0)
})
```


Heatmap {data-icon="fa-th"}
=====================================

### Association Heatmap {.no-title}

```{r}
plotlyOutput("heatmap_main", height = "100%")

output$heatmap_main <- renderPlotly({ plot_heatmap(fd()) })
```


BMI Comparison {data-icon="fa-balance-scale"}
=====================================

Column {data-width=50%}
-------------------------------------

### BMI-Adjusted heatmap

```{r}
plotlyOutput("hmap_bmi", height = "100%")
output$hmap_bmi <- renderPlotly({
  sub <- dat_bmi %>%
    dplyr::semi_join(fj(), by = c("gene_snp_ra","is_index"))
  plot_heatmap(sub, min_h = 420)
})
```

### BMI-Unadjusted heatmap

```{r}
plotlyOutput("hmap_nobmi", height = "100%")
output$hmap_nobmi <- renderPlotly({
  sub <- dat_no_bmi %>%
    dplyr::semi_join(fj(), by = c("gene_snp_ra","is_index"))
  plot_heatmap(sub, min_h = 420)
})
```

Column {data-width=50%}
-------------------------------------

### ╬ö╬▓ map  (no-BMI ÔêÆ BMI-adj)  ÔÇö purple = BMI attenuates, green = BMI amplifies

```{r}
plotlyOutput("hmap_diff", height = "100%")
output$hmap_diff <- renderPlotly({ plot_bmi_diff_heatmap(fj(), min_h = 420) })
```

### ╬▓ scatter  (BMI-adj vs no-BMI)

```{r}
plotlyOutput("scatter_bmi", height = "100%")
output$scatter_bmi <- renderPlotly({ plot_beta_scatter(fj()) })
```


Volcano {data-icon="fa-chart-line"}
=====================================

Column {.sidebar data-width=220}
-------------------------------------

```{r}
selectInput("volcano_idx", "IS Index",
  choices  = c("All Indices", IS_INDEX_ORDER),
  selected = "All Indices",
  width    = "100%"
)

p("Each dashed line marks a significance threshold (orange: p=0.05, red: p=0.001).",
  style = "font-size:0.75rem; color:#777; margin-top:0.5rem")
```

Column
-------------------------------------

### Volcano Plot {.no-title}

```{r}
plotlyOutput("volcano_main", height = "100%")
output$volcano_main <- renderPlotly({ plot_volcano(fd(), input$volcano_idx) })
```


Effect Profile {data-icon="fa-bullseye"}
=====================================

Column {.sidebar data-width=260}
-------------------------------------

```{r}
renderUI({
  choices <- c("" = "", sort(unique(fd()$gene_snp_ra)))
  selectizeInput("forest_var", "Select variant (gene ÔÇô rsID ÔÇô allele)",
    choices  = choices,
    selected = "",
    options  = list(placeholder = "Type to searchÔÇª"),
    width    = "100%"
  )
})

p("Shows ╬▓ across all 21 IS indices for the selected variant.
  Both analyses are overlaid for direct comparison.",
  style = "font-size:0.75rem; color:#777; margin-top:0.5rem")
```

Column
-------------------------------------

### Effect Profile ÔÇö both analyses overlaid {.no-title}

```{r}
plotlyOutput("forest_main", height = "100%")
output$forest_main <- renderPlotly({
  var <- input$forest_var %||% ""
  if (nchar(trimws(var)) == 0)
    return(.empty_plot("Select a variant in the sidebar dropdown"))

  grps <- input$groups %||% c("Fasting","OGTT,0-120","OGTT,0-30-120")
  plot_forest(
    dat_bmi    %>% dplyr::filter(index_group %in% grps),
    dat_no_bmi %>% dplyr::filter(index_group %in% grps),
    var
  )
})
```


BMI Mediation {data-icon="fa-arrows-alt-h"}
=====================================

Column {data-width=50%}
-------------------------------------

### % ╬▓ mediated by BMI ÔÇö top variants

```{r}
plotlyOutput("attenuation_bar", height = "100%")
output$attenuation_bar <- renderPlotly({ plot_attenuation_bar(fj()) })
```

Column {data-width=50%}
-------------------------------------

### Parallel coordinates ÔÇö all variants ├ù IS indices

```{r}
plotlyOutput("parallel_main", height = "100%")
output$parallel_main <- renderPlotly({ plot_parallel(fd()) })
```


Data Browser {data-icon="fa-table"}
=====================================

Column {.sidebar data-width=240}
-------------------------------------

```{r}
radioButtons("table_view", "Show data for",
  choices  = c("BMI-Adjusted"  = "bmi",
               "BMI-Unadjusted"= "no_bmi",
               "Both (joined)" = "joined"),
  selected = "bmi"
)

checkboxInput("sig_only", "Significant only (p < 0.05)", FALSE)

hr()

downloadButton("dl_csv", "Ô¼ç  Download CSV",
               class = "btn btn-outline-primary btn-sm w-100")

p("Use column headers to filter rows. Click any column header to sort.",
  style = "font-size:0.75rem; color:#777; margin-top:0.8rem")
```

Column
-------------------------------------

### Data Table {.no-title}

```{r}
tbl_data <- reactive({
  if (input$table_view == "joined") {
    base <- fj() %>%
      dplyr::select(nearest_gene, index_variant, t_2_d_risk_allele,
                    gene_snp_ra, is_index, index_group,
                    beta_bmi, pval_bmi, beta_no_bmi, pval_no_bmi)
    if (input$sig_only)
      base <- dplyr::filter(base, pval_bmi < 0.05 | pval_no_bmi < 0.05)
  } else {
    base <- fd()
    if (input$sig_only)
      base <- dplyr::filter(base, pvalue < 0.05)
  }
  base
})

output$dl_csv <- downloadHandler(
  filename = function() paste0("isi_t2d_", input$table_view, "_", Sys.Date(), ".csv"),
  content  = function(file) readr::write_csv(tbl_data(), file)
)

DT::renderDT({
  df <- tbl_data()
  beta_cols <- intersect(c("beta","beta_bmi","beta_no_bmi"), names(df))
  pval_cols <- intersect(c("pvalue","pval_bmi","pval_no_bmi"), names(df))

  DT::datatable(
    df,
    filter     = "top",
    rownames   = FALSE,
    extensions = c("Buttons","Scroller"),
    options    = list(
      dom         = "Bfrtip",
      buttons     = list(
        list(extend = "csv",    text = "CSV"),
        list(extend = "excel",  text = "Excel"),
        list(extend = "colvis", text = "Columns")
      ),
      scrollY     = "calc(100vh - 280px)",
      scroller    = TRUE,
      deferRender = TRUE,
      pageLength  = 50,
      columnDefs  = list(list(className = "dt-center", targets = "_all"))
    )
  ) %>%
    DT::formatRound(beta_cols, digits = 5) %>%
    DT::formatSignif(pval_cols, digits = 3) %>%
    DT::formatStyle(
      pval_cols,
      background = DT::styleInterval(
        c(0.001, 0.05),
        c("#f8d7da80", "#fff3cd80", "transparent")
      ),
      fontWeight = DT::styleInterval(0.05, c("bold", "normal"))
    )
}, server = TRUE)
```


Summary {data-icon="fa-chart-bar"}
=====================================

Column {data-width=55%}
-------------------------------------

### Top genes by number of significant associations (p < 0.05)

```{r}
plotlyOutput("gene_bar", height = "100%")
output$gene_bar <- renderPlotly({ plot_gene_summary(fd()) })
```

Column {data-width=45%}
-------------------------------------

### ╬▓ distribution per IS index

```{r}
plotlyOutput("index_violin", height = "100%")
output$index_violin <- renderPlotly({ plot_index_violin(fd()) })
```
